---
title: "Memory-bounded NRV summaries with nrvtools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Memory-bounded NRV summaries with nrvtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup}
library(nrvtools)
```

## The workflow

A Natural Range of Variation (NRV) analysis runs the same landscape many times
(replicates) and characterises how a set of metrics varies across those
replicates over time.
With large study areas and many replicates the raw per-replicate metric tables
do not fit comfortably in memory all at once, so nrvtools keeps the reduction
memory-bounded:

1. compute the **raw** per-replicate metrics
   (`nrv_metrics_landscape()`, `calculatePatchMetrics()`,
   `calculatePatchMetricsSeral()`);
2. tidy each replicate's metrics into one long table (`tidy_nrv_metrics()`) and
   write it to its own partition of an on-disk Arrow dataset
   (`write_nrv_parquet()`);
3. open the dataset lazily (`open_nrv_dataset()`) and reduce across replicates
   by pushing the aggregation down to Arrow compute (`summarize_nrv()`), so the
   per-replicate rows are never all held in memory at once;
4. plot the resulting range-of-variation envelope (`plot_nrv_envelope()`).

Every raw producer emits the same long schema -- `level`, `class`, `metric`,
`value`, plus `rep`, `time`, and `poly` -- so a single reducer handles landscape-
and patch-level metrics alike.
The \pkg{landscapemetrics} `metric` name together with `level` identifies each
metric, so no metric-name book-keeping columns are needed.

## A raw metric producer

The raw producers read maps from disk and parse `rep`/`time`/`poly` from the
file paths.
For a self-contained example we call `patchAreas()` -- the per-patch area
calculator that `calculatePatchMetrics()` wraps -- directly on a small in-memory
categorical raster:

```{r mock-raster}
library(terra)

make_vtm <- function(seed) {
  set.seed(seed)
  m <- matrix(sample(1:3, 25, replace = TRUE), nrow = 5)
  r <- terra::rast(m, crs = "EPSG:3857")
  levels(r) <- data.frame(ID = 1:3, values = c("conifer", "mixed", "deciduous"))
  r
}

patchAreas(make_vtm(1))
```

Each row is one patch: the raw, unaggregated values that feed the summary.

## Build a replicate dataset

In a real analysis each replicate's tidied metrics are written to its own
partition as the run completes; here we loop over a few mock replicates and two
time steps, stamping `time`/`poly` the way the producers would:

```{r build-dataset}
root <- file.path(tempdir(), "nrv-demo")
unlink(root, recursive = TRUE)

for (rep_id in 1:5) {
  per_time <- lapply(c(0L, 100L), function(t) {
    areas <- patchAreas(make_vtm(rep_id * 10 + t))
    areas$time <- t
    areas$poly <- "unit-A"
    areas
  })
  ## tidy_nrv_metrics() row-binds a raw metric list into one long table
  tidied <- tidy_nrv_metrics(per_time, studyArea = "demo", scenario = "hrv")
  write_nrv_parquet(tidied, root, replicate = rep_id)
}

list.files(root, recursive = TRUE)[1:3]
```

Each replicate is an isolated `replicate=<id>/part-0.parquet` partition, so many
replicate writers can run concurrently -- even against an NFS mount -- without
colliding.

## Summarize across replicates

`summarize_nrv()` opens the dataset lazily and returns the across-replicate
envelope (replicate count, mean, sd, min, max, median, plus standard error and a
95% confidence half-width) for each combination of the identifier columns:

```{r summarize}
env <- summarize_nrv(root)
env
```

Only this small envelope is ever materialised in R; the per-replicate patch rows
stay on disk.

## Plot the envelope

`plot_nrv_envelope()` draws the across-replicate mean with a min--max ribbon,
faceting by whichever categorical columns vary:

```{r plot}
plot_nrv_envelope(env, ylab = "patch area (ha)")
```

For the paginated, faceted-by-reporting-unit plots used in SpaDES
post-processing, see `plot_over_time()`, `plot_over_time_by_class()`, and
`plot_by_class()`, which consume the same `summarize_nrv()` envelope.
Order seral-stage classes for display with
`factor(class, levels = seral_stages())`.
